Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-11-06 paper2_JNLS/paper.tex
NM

Rogue waves are abnormally large surface waves, defined by oceanographers as those that exceed twice the significant wave height \citep{muller2005rogue, ying2011linear}. Though such waves were once dismissed as myth, they have now been recorded in oceans across the globe and pose a recognized threat to seagoing vessels and naval structures. Rogue waves, also variously known as freak or anomalous waves, have been observed both in deep and shallow water, and can be triggered by a variety of mechanisms, including anomalous wind-forcing \citep{kharif2008influence, toffoli2017wind}, opposing currents \citep{garrett2009rogue, onorato2011triggering},
focusing due to variable bathymetry \citep{heller2008refraction, white1998chance}, 
and the Benjamin-Feir deep-water modulational instability \citep{benjamin1967disintegration, viotti2013emergence, cousins2015unsteady, farazmand2017reduced}.
The common tie between these mechanisms is their ability to generate non-normal statistics in the surface displacement; when governed by Gaussian statistics, the likelihood of a rogue wave is extremely low, but these rare events occur much more frequently when surface statistics deviate from Gaussian. In this way, it makes sense to approach anomalous waves from the broader perspective of turbulent dynamical systems \cite{sapsis2013a, sapsis2013b, sapsis2013blending, chen2016filtering, majda2016introduction, macedo2017universality, MajdaQiSIAM2018, blonigan2019extreme, guth2019machine}.

@@

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda and Qi analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure \cite{majdaqi2019}.

@@

@@

% Figure: Experimental Spatial Statistics
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.80 \linewidth]{Figs/ExpSpatialStats.pdf}
\caption{Wave statistics as they vary in space for several experiments of different driving amplitudes (see legend). (a)--(c) Standard deviation, skewness, and kurtosis of the surface displacement $\eta$. (d)--(f) The same for surface slope $\eta_x$. (a) $\etastd$, which sets a scale for wave amplitude, does not vary significantly crossing the ADC. (b)--(c) Skewness and kurtosis of $\eta$, however, show a strong response to the ADC. (d)--(f) The measurements of surface slope, $\eta_x$, are negatively skewed and also exhibit large kurtosis downstream of the ADC. }
\label{ExpSpatialStats}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

	We now aim to compare the wave statistics measured in experiments against those that emerge from the TKdV theory. Throughout, we will focus on the representative set of experiments detailed in Table \ref{paramtable}, in which the peak downstream skewness was measured to be 0.83. 
%We can then use Fig.~\ref{transfig}(c) as a guide to select the appropriate upstream inverse temperature $\thup$. With $(\Lambda, \lamfac) = (16, 8)$ fixed, the figure indicates a value of $\thup = 17$ achieves the desired outgoing skewness. We thus run the deterministic TKdV simulations with model parameters $\Lambda = 16$, $\lamfac = 8$, $\thup = 17$ and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. 
%
With $(\Lambda, \lamfac) = (16, 8)$ fixed, Fig.~\ref{transfig}(c) indicates $\thup = 20$ as a reasonable value to attain the desired skewness. We thus run the deterministic TKdV simulations with these model parameters and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. We run the simulations to a sufficiently long dimensionless time of $t_{f} = 10$, with $dt = 5 \times 10^{-4}$ and with $10^3$ trajectories sampled from the upstream Gibbs measure $\Gup$.

@@

	The comparison between experiments and theory in Fig.~\ref{uhist} is striking. As seen in the figure, both the experiments and theory show a transition from a nearly symmetric upstream distribution, to a highly skewed distribution downstream. The TKdV theory not only captures this transition, but also recovers the shape of the resulting downstream distribution in remarkable detail. Nearly every feature that can be compared --- the decay rate of the tail, the position and value of the peak, the rapid cutoff for negative $u$ --- matches surprisingly well. This comparison offers compelling visual evidence for: (a) the predictive power of the TKdV framework, and (b) the successful calibration of the incoming inverse temperature.

@@

% Figure
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.99 \linewidth]{Figs/slopehist.pdf}
\caption{
Comparison of slope histograms. The TKdV simulations show similar slope distributions upstream, but visual distributions in the downstream distributions are visible. While the simulations recover the spread of the downstream slope statistics, they do not capture the asymmetry seen in the experiments.
}
\label{slopehist}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

	Figure \ref{slopehist} shows the distributions of surface slope that result from experiments (top) and theory (bottom). These histograms show intriguing similarities and differences. First, the upstream slope distribution is captured well by the TKdV simulations, both in its nearly symmetric shape and its scale. We note that, while the standard deviation of displacement was input into the TKdV theory, the slope standard deviation was not. Downstream, both the experimental and theoretical slope distributions  spread significantly. Interestingly, in the experiments, it is not the standard deviation that increases significantly upon crossing the depth change ($\std(\eta_x)$ increases from 0.15 to 0.17), but instead the excess kurtosis, which jumps from a value of 0.46 upstream to 3.4 downstream. Likewise, in the TKdV theory, $\std(\eta_x)$ grows from 0.13 upstream to 0.33 downstream, and $\kurt(\eta_x)$ grows from -0.04 to 0.50. Thus, the theoretically predicted jump in kurtosis, while not as extreme as that measured in experiments, is quite significant. These elevated levels of kurtosis are associated with the distinct appearance of the distributions, most notably the flat tails that appear in both the experimental and theoretical distributions, and also the pointy center that is most pronounced in the experiments.

	Distinct differences in the shapes of these downstream distributions, however, are also visible. While the theoretical distributions remain symmetric, the experimental distributions skew towards negative slope. As mentioned earlier, this skewness is consistent with a right-moving wave of steep leading surface (i.e.~negative slope), which is a phenomenon apparently not captured by the TKdV simulations.
Finally, it should be commented that the experimental measurements of slope involve numerical differentiation of the free surface extracted from optical images, which naturally introduces error. Furthermore, high-order statistics, such as kurtosis, can be very sensitive to noise and measurement error. Hence, the experimental statistics of slope, while highly valuable, should be regarded with some level of caution.

@@

These observations can be rationalized with some simple analysis of the structure of TKdV. Upstream of the ADC, since $\drat = 1$, the nonlinear effects are relatively weak and the dynamics are dominated by dispersion. Linearizing \eqref{TKdV} and introducing the ansatz $u_k = e^{i(kx - \omega_k t)}$ produces the dispersion relation for angular frequency $\omega_k = -C_2 k^3$, or for phase velocity $c_k = \omega_k/k = -C_2 k^2$ \cite{majdaqi2019}. In particular, the phase velocity is strictly negative, implying that waves can only propagate leftwards, as is consistent with Fig.~\ref{trajectories}(a).
% Further details, see page 6 of Majda Qi, J Stat Phys.
%$C_3 = 0.20$ and $C_2 = 0.024$.

	Downstream of the ADC, however, $\drat = 0.24$, which significantly amplifies nonlinearity while suppressing dispersion, thereby allowing waves to propagate in either direction. In particular, the dispersionless limit of KdV is the Burgers-Hopf equation, in which the nonlinear advection is proportional in both amplitude and direction to the value of $u$. Hence, positive displacements would be expected to move rightwards, as is seen in Fig.~\ref{trajectories}(b).

>>>>

Rogue waves are abnormally large surface waves, defined by oceanographers as those that exceed twice the significant wave height \citep{muller2005rogue, ying2011linear}. Though such waves were once dismissed as myth, they have now been recorded in oceans across the globe and pose a recognized threat to seagoing vessels and naval structures. Rogue waves, also variously known as freak or anomalous waves, have been observed both in deep and shallow water, and can be triggered by a variety of mechanisms, including anomalous wind-forcing \citep{kharif2008influence, toffoli2017wind}, opposing currents \citep{garrett2009rogue, onorato2011triggering},
focusing due to variable bathymetry \citep{heller2008refraction, white1998chance}, 
and the Benjamin-Feir deep-water modulational instability \citep{benjamin1967disintegration, viotti2013emergence, cousins2015unsteady, farazmand2017reduced}.
The common tie between these mechanisms is their ability to generate non-normal statistics in the surface displacement; when governed by Gaussian statistics, the likelihood of a rogue wave is extremely low, but these events occur much more frequently when surface statistics deviate from Gaussian. In this way, anomalous waves can be approached from the broader perspective of turbulent dynamical systems \cite{sapsis2013a, sapsis2013b, sapsis2013blending, chen2016filtering, majda2016introduction, macedo2017universality, MajdaQiSIAM2018, blonigan2019extreme, guth2019machine}.

@@

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda and Qi analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure \cite{majdaqi2019}. More recently, Qi and Majda demonstrated the capability of machine learning strategies to accurately predict these extreme events \cite{MachineLearning2019}. That work uses a deep neural network coupled with a judicious choice of the entropy loss function that emphasizes the dominant structures of the turbulent field \cite{MachineLearning2019}. It should be noted that, in these studies, the extreme events can occur intermittently and on a relatively frequent basis, in contrast to the simpler situation of isolated, rare events \cite{guth2019machine}.

@@

% Figure: Experimental Spatial Statistics
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.8 \linewidth]{Figs/ExpSpatialStats.pdf}
\caption{Wave statistics as they vary in space for several experiments of different driving amplitudes (see legend). (a)--(c) Standard deviation, skewness, and kurtosis of the surface displacement $\eta$. (d)--(f) The same for surface slope $\eta_x$. (a) $\etastd$, which sets a scale for wave amplitude, does not vary significantly crossing the ADC. (b)--(c) Skewness and kurtosis of $\eta$, however, show a strong response to the ADC. (d)--(f) The measurements of surface slope, $\eta_x$, are negatively skewed and also exhibit large kurtosis downstream of the ADC. }
\label{ExpSpatialStats}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

@@

	We now aim to compare the wave statistics measured in experiments against those that emerge from the TKdV theory. Throughout, we will focus on the representative set of experiments detailed in Table \ref{paramtable}, in which the peak downstream skewness was measured to be 0.83. 
With $(\Lambda, \lamfac) = (16, 8)$ fixed, Fig.~\ref{transfig}(c) indicates $\thup = 20$ as a reasonable value to attain the desired skewness. We thus run the deterministic TKdV simulations with these model parameters and with $\epsup$, $\delup$, and $\dratdn$ set to the representative values in Table \ref{paramtable}. We run the simulations to a sufficiently long dimensionless time of $t_{f} = 10$, with $dt = 5 \times 10^{-4}$ and with $10^3$ trajectories sampled from the upstream Gibbs measure $\Gup$.

@@

	The comparison between experiments and theory in Fig.~\ref{uhist} is striking. As seen in the figure, both the experiments and theory show a transition from a nearly symmetric upstream distribution to a highly skewed distribution downstream. The TKdV theory not only captures this transition, but also recovers the shape of the resulting downstream distribution in remarkable detail. Nearly every feature that can be compared --- the decay rate of the tail, the position and value of the peak, the rapid cutoff for negative $u$ --- matches surprisingly well. This comparison offers compelling visual evidence for: (a) the predictive power of the TKdV framework, and (b) the successful calibration of the incoming inverse temperature.

@@

% Figure
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.99 \linewidth]{Figs/slopehist.pdf}
\caption{
Comparison of surface slope histograms. The experiments and TKdV simulations show very similar slope distributions upstream. Downstream, the theory accounts for the spread of the distribution and the long exponential tails, while differences in the detailed shape of the distributions are also visible. Note the elevated uncertainty in the experimental data due to numerical differentiation of free-surface measurements and the associated amplification of noise.}
\label{slopehist}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

	Figure \ref{slopehist} shows the distributions of surface slope that result from experiments (top) and theory (bottom). These histograms show intriguing similarities and differences. First, the upstream slope distribution is captured well by the TKdV simulations, both in its nearly symmetric shape and its scale. We note that, while the standard deviation of displacement was input into the TKdV theory, the slope standard deviation was not. Downstream, both the experimental and theoretical slope distributions  spread significantly. Interestingly, in the experiments, it is not the standard deviation that increases significantly upon crossing the depth change ($\std(\eta_x)$ increases from 0.15 to 0.17), but instead the excess kurtosis, which jumps from a value of 0.46 upstream to 3.4 downstream. Likewise, in the TKdV theory, $\std(\eta_x)$ grows from 0.13 upstream to 0.33 downstream, and $\kurt(\eta_x)$ grows from -0.04 to 0.50. Thus, the theoretically predicted jump in kurtosis, though not as extreme as that measured in experiments, is quite significant. These elevated levels of kurtosis are associated with the distinct appearance of the downstream distributions, most notably the long, flat tails that appear in both experiments and theory.

	Differences in the detailed shapes of these downstream distributions are also visible. First, we point out that the experimental measurements involve numerical differentiation of  surface displacement extracted from optical images, a process that unavoidably amplifies any noise that is present. We must therefore proceed with caution in comparing the experiment and theory, recognizing the possibility that observed discrepancies may be due to these measurement errors. Nonetheless, we notice that the theoretically predicted distributions remain symmetric downstream and the experimental ones skew towards negative slope. As mentioned in Section \ref{experiments}, negative skewness is consistent with a right-moving wave of steep leading surface (i.e.~negative slope). Additionally, the peak of the experimental distribution appears sharper than that of the theory.

%Figure \ref{ExpSpatialStats} may offer additional insight on these detailed differences in the shape of the downstream slope distributions. In particular, \ref{ExpSpatialStats}(e)--(f) show the spatial variation of the slope skewness and kurtosis. Note that variations in these statistics are significantly more transient than the corresponding variations in surface displacement; i.e.~the peaks in surface slope statistics arise more rapidly and decay more rapidly. Variations in displacement statistics also decay, but over a longer spatial scale, presumably due to dissipative effects. The TKdV histograms shown in Fig.~\ref{slopehist}, meanwhile, are long-time averages and hence not intended to capture such transient effects. These observations offer the suggestion that slope dynamics experience stronger transients than do surface displacements, leading to greater difficulty in the direct comparison between measurements extracted from a single location (experiments) and long-time averages (theory).

@@

These observations can be rationalized with some simple observations regarding the structure of the TKdV system. Upstream of the ADC, since $\drat = 1$, the nonlinear effects are relatively weak and the dynamics are dominated by dispersion. Linearizing \eqref{TKdV} and introducing the ansatz $u_k = e^{i(kx - \omega_k t)}$ produces the dispersion relation for angular frequency $\omega_k = -C_2 k^3$, or for phase velocity $c_k = \omega_k/k = -C_2 k^2$ \cite{majdaqi2019}. In particular, the phase velocity is strictly negative, implying that waves only propagate leftwards, as is consistent with Fig.~\ref{trajectories}(a).
% Further details, see page 6 of Majda Qi, J Stat Phys.
%$C_3 = 0.20$ and $C_2 = 0.024$.

	Downstream of the ADC, however, the depth ratio changes to $\drat = 0.24$. This change significantly amplifies nonlinearity while suppressing dispersion, and thereby allows waves to propagate in either direction. In particular, the dispersionless limit of KdV is the Burgers-Hopf equation, for which the nonlinear advection is proportional in both amplitude and direction to the value of $u$. Hence, positive displacements would be expected to move rightwards, as is seen in Fig.~\ref{trajectories}(b).
